---
title: "W11L2: Causality"
subtitle: "POLS 642 Intermediate Analysis of Political Data"
author: "Dr. Ches Thurber"
format: 
  revealjs:
    theme: simple
execute: 
  echo: true
editor: visual
---

## Plan for Today

1\) "Bell Ringer"

2\) Causality

3\) Experiments

4\) Observational Studies

## Packages

```{r}
#install.packages("ggdag")
library(ggdag)
```

## Bell-Ringer

I have really cool data on campaign spending in municipal elections in Indonesia. My theoretical intuition is that greater spending will cause candidates to get higher vote shares. I run a bivariate regression and... STARS! But...

1\) Dr. Schrauf says I need to "control" for candidate quality. Should I?

2\) My friend says I need to control for whether the candidate spent money on radio advertising. Should I?

3\) If I still have a statistically significant result, can I claim that I have shown causation?

# Causality

## Defining Causation

Multiple philosophical definitions:

1\) X is associated with Y, temporally precedes it, with no supurriousness (correlational)

2\) Y "listens" to X, and behaves in response (mechanistic)

3\) Absent X, Y would be different (counterfactual)

4\) It's all just a social construct, man. Better just to interrogate your positionality...(constitutive)

## The Problem of Endogeneity

-   in order to infer causality, we must assume that the treatment and control groups are identical in all ways that might affect the outcome

-   *endogeneity* refers to any theoretical process that might make this assumption untrue. It usually takes one of two forms:

-   Reverse causality: Y causes X

-   Omitted Variable Bias (OVB) or a confounding variable: a third variable Z causes both X and Y

## Example

-   as publicly-funded but privately-run "charter schools" became widespread in the United States, policymakers and citizens wanted to know whether they produced better outcomes than public schools

-   First, analysts compared the test scores of students in public schools versus charter schools. They were higher in charter schools!

-   Charter schools were associated with higher test scores. But did they *cause* those higher test scores?

## Confounding Variables

A confounding variable is any variable that:

1\) affects the likelihood of receiving treatment $X$ , AND

2\) affects the outcome $Y$

```{r, echo = F}
ggdag_confounder_triangle() + theme_dag()
```

## The Confounder

-   in the case of charter schools, greater *family income* might make students more likely to choose a charter school AND more likely to achieve higher test scores

    ```{r, echo = F}

    ggdag_confounder_triangle() + theme_dag()
    ```

## The Problem

-   our treatment and control groups are not identical in a way that likely affects test scores

-   the higher average test scores for charter school students may be the result of the treatment group's higher average family income, rather than the result of charter schools

-   our results are biased because of an omitted confounding variable: family income

## The Solutions

1\) Experimental research designs (Randomized Controlled Trials or RCTs)

2\) Multivariate Regression: adding additional variables to a regression as "controls"

3\) Quasi-experimental approaches (instrumental variables, difference-in-differences, regression discontinuity designs, matching methods)

# Experiments

## Experiments: The Beauty of Randomization

-   when you are able to randomize treatment, you know you should have two groups – treatment and control – that should be, on average, identical in all observed and unobserved pre-treatment characteristics

-   that is, you can assume that the treated and untreated observations are similar with respect to all the variables that might affect the outcome other than the treatment itself

## Charter School "Lotteries"

-   Researchers looked at charter schools that had held random lotteries to determine who got in.

-   Comparing the students who "won" the lottery and entered the charter school to those who "lost" the lottery and went to public school

-   The lottery created *randomized treatment*, ensuring that the treatment and control groups were identical

-   Most studies using this approach have shown no causal effects for charter schools

## Drawbacks of Experiments

-   Ethics: is it fair to deny some people treatment? or to apply a treatment when the result is unknown and potentially dangerous?

-   Logistics: some things just can't be randomly assigned

-   Financial: experiments are expensive

So if we can't do an experiment, what can we do instead?...

# Causal Inference in Observational Studies

## Multiple Linear Regression

1\) Identify all potential confounding variables

2\) Add potential confounders as additional $X$ variables in a regression model, and model them correctly (transformations, etc...)

## Limitations

-   can we ever truly know that we have included all potential confounders or that we have modeled them correctly??

-   don't want to include *post-treatment* variables: those that might be caused by our treatment

## Mediators

-   these are intervening variables, part of HOW X causes Y

-   e.g. teacher quality or hours studied might be mediators in our charter school example

-   We do NOT want to control for these

```{r, echo = F}
ggdag_mediation_triangle() + theme_dag()
```

## Colliders

-   variables that are caused by both X and Y

-   college acceptance would be a collider in our charter school example

-   We do NOT want to control for these

    ```{r, echo = F}
    ggdag_collider_triangle() + theme_dag()
    ```

# Let's Map Out Some Cases

## Causation and Plain Regression

-   Multivariate regression is one way of moving toward causation

-   But it requires some heroic assumptions to truly capture a causal relationship

-   as such we never use causal language

-   instead use the language of comparison

## Language of Comparison

"In our study, candidates who had \$1 more in funding achieved on average a $\hat{\beta_1}$ greater share of the vote, holding other factors constant"

## Best Practices in large-N observational work

-   be careful with the language you use

-   clearly specify potentially confounding variables and cite sources to back up

-   maybe even include causal diagram

-   explicitly discuss risks and sources of omitted variable bias

-   sensitivity analysis

## Toward causation in observational studies

-   Other techniques might allow us to relax some of these assumptions

-   Matching methods do not require perfect linearity of confounding variables

-   Fixed Effects will allow us to control for whole sets of possible confounding variables that are in turn correlated with groups in our data

-   Instrumental variables will allow us to find plausibly random variation in our independent variables

## 
