---
title: "W3L2: Comparing Variables"
subtitle: "POLS 642 Intermediate Analysis of Political Data"
author: "Dr. Ches Thurber"
format: 
  revealjs:
    theme: simple
execute: 
  echo: true
editor: visual
---

## Get Started

1\) Follow link on Blackboard to download `W3L2_Comparing_Variables.zip`

2\) "Unzip" it and get the files set up in your file directory the way you want them

3\) Open up your R Project

## Big Picture

-   Today: Describing Two Variables

-   Fri: Data Camp Due (Probability, Hypothesis Testing)

-   Tue: Fitting Regressions

-   Thur: Hypothesis Testing Regressions

-   Fri: More Data Camp

## Plan for Today

1\) Recap Hyp Testing and Conf Intervals

2\) Means and Normal Distributions

3\) T-Distributions

3\) Comparing Two Variables

# Recap

## Descriptive Inference

-   Descriptive Inference: estimate the characteristics of the *population* based on the *sample*

-   We have in our hands a *sample*, we can calculate descriptive statistics, plot its distribution, etc...

-   But what we really want to know about is the *population*

## Core Problem

-   we can know the probability of a model/distribution producing a certain outcome
-   we cannot know the probability of a model being true given an outcome

## What to do?

"One thing that is a bit easier to do is to learn *whether certain theoretical distributions are unlikely.* Maybe we can’t figure out exactly what the theoretical distribution *is*, but we can rule some stuff out."

-   *The Effect*

## What to do?

-   Hypothesis Testing approach: Set a "null hypothesis" about the population parameter and ask: "If the null hypothesis were true, how likely is my sample?"

-   Confidence Intervals: Identify the range of population parameters from which drawing your sample would be reasonably likely

    They are both mathematically the same: If the null hypothesis is not within the confidence interval, you would reject the null!

## Hypothesis Testing approach

-   establish **null hypothesis** of what a core characteristic of the theoretical distribution of the population might be (e.g. its mean or its $\pi$)

-   calculate probability of drawing a sample from the null hypothesis's theoretical distribution with core characteristic of your observed sample

-   reject the null hypothesis when the probability of drawing a sample with the observed characteristic as extreme as yours is small (often $\alpha = 0.05$)

## Hypothesis testing the Coin flip

-   **Null Hypothesis**: The coin is "fair" (i.e. the population distribution is binomial with $\pi$ = 0.5)

-   **Alternative Hypothesis**: The coin is not fair (i.e. $\pi \neq .5$)

```{r, six flips}
dbinom(x=6, size=6, prob=0.5)
```

-   Because .016 \< .05, we reject the null, and accept the alternative hypothesis: Thurber is a liar!

# Polling

## Let's Look at Today's Rasmussen Poll

-   1500 respondents

-   47% approve of Trump

-   52% disapprove

## Calculating the Test

What's the probability of getting an approval rating of 47% or less (or 53% or more) if the real approval rating were 50%, with a sample size of 1500 under the conditions of the null?

```{r}
2 * pbinom(.47*1500, 1500, .5)
```

Can we reject the null?

## A Different Way

-   What if, instead of asking whether or not our sample is a reasonably likely outcome from one specific theoretical distribution, we asked, "What is the range of theoretical distributions from which our sample is a reasonably likely outcome?"

-   To put it differently, "what is the set of null hypotheses that we would fail to reject?"

-   What are the values of $\pi$ for which our sample statistic falls within the 95% of most probably outcomes

## Calculating the Bounds

What is the lowest actual Trump approval rating for which 47% still has a 2.5% chance of occurring?

What is the highest actual Trump approval rating for which 46% still has a 2.5% chance of occurring?

```{r}

qbinom(.025, 1500, .47) /1500

qbinom(.975, 1500, .47) /1500
```

Our sample of 47% approval is reasonably likely if the actual population $\pi$ of people who support Trump is anywhere between 44.5% and 49.5% (i.e. $\pm$ 2.5%)

## 

# Continuous Variables and Normal Distributions

## Back to BES

```{r}
BES <- na.omit(read.csv("BES.csv")) #never do this
head(BES)
```

## Distribution of Ages

Is this normal?!?

```{r}
hist(BES$age)
```

## Distribution of Sample Means

```{r}
sample_means <- replicate(1000,
                          mean(sample(BES$age, size = 1000, replace = F), na.rm=T))
hist(sample_means)
```

## Central Limit Theorem

As you take more samples, the distribution of the sample means converges on a normal disribution with:

-   a mean that is the same as that of the population ($\mu$)

-   a standard deviation (standard error) that is that of the population divided by the square root of the size of the samples:

    $$
    \frac{\sigma}{\sqrt(n)}
    $$

-   this is true no matter the underlying distribution

## Standard Deviation vs. Error

-   a **standard deviation** is a parameter of a population or a statistic we can calculate from a sample that measures the dispersion of a variable from the mean

-   the **standard error** is the the standard deviation of the distribution of sample means when drawing repeated samples from a population

    -   the standard error is always smaller than the standard deviation

    -   its size depends on the standard deviation of the population and the size of the samples we are drawing

## Sample Mean and Sample Std. Dev

What's the average age of the British voter?

```{r}
mean(BES$age)
mean(sample_means)
sd(BES$age)
sd(BES$age)/sqrt(100)
sd(sample_means)

```

## What about the population of British Voters?

-   let's take our whole BES dataset as the sample

-   the population is all British voters

-   could the average age of all British voters be 50.5?

## Let's Look at the Normal Distribution

![](Normal.svg)

## What we know from the Normal Distribution

-   we know that 95% of **sample means** will fall within about two **standard errors** of the population mean

-   So if the population mean $\mu$ were 50.5, then we would expect 95% of surveys of 25097 British voters to fall between two standard errors of 50.5

-   what's the standard error for our survey of 25097?

    ```{r}
    std.error <- sd(BES$age)/sqrt(25097)
    std.error
    ```

## Interpretation

-   If the population mean were 50.5, we would expect 95% of surveys of 25k voters to produce sample mean ages between 50.3 and 50.7

-   Our sample mean of 50.64 is within that range, so we conclude that 50.5 is possibly the average age of the population of British voters

## Calculating the p-value

-   We could formally calculate the probability of getting a sample as **extreme** as ours (50.64 or higher, or 50.36 or lower):

    ```{r}
    left.tail <- pnorm(50.36, mean= 50.5, sd= .1)
    left.tail
    right.tail <- 1-pnorm(50.64, mean= 50.5, sd= .1)
    right.tail

    left.tail + right.tail

    ```

-   We also know that the confidence interval, that is the set of population means for which our sample mean falls within two standard errors is 50.64 $\pm$ 0.2, or 50.44-50.84

# The T-Distribution

## A catch

-   Wait, wait, wait!

-   The standard error formula contains $\sigma$ which is the standard deviation of the population

-   I don't actually know this! (Just like I don't actually know the population mean)

-   I cheated and plugged the sample standard deviation in for the population standard deviation

## What to do instead

-   actually, this is an okay assumption to make! BUT, it requires that we use the t-distribution rather than the normal distribution

-   the t-distribution is a lot like the normal distribution, it just has "fatter tails" for smaller sample sizes

-   it accounts for the fact that our sample standard deviation is more likely to vary from the population when our sample size is small

-   the "two standard errors away from the mean" rule of thumb is still pretty accurate

## A formal calculation with the t-distribution

```{r}

left.t.tail <- pt((50.5-50.64)/.1, df = 25097)
left.t.tail
left.tail

2* left.t.tail
```

# Comparing Two Variables

## Conditional Means

-   what if we wanted to know the average age of voters who intend to vote stay? who intend to vote leave?

-   this is called a conditional mean

    ```{r}
    mean.stay <- mean(BES$age[BES$vote=="stay"])
    mean.leave <- mean(BES$age[BES$vote == "leave"])
    mean.stay
    mean.leave
    ```

## Visually

```{r, echo = F}
library(ggplot2)
ggplot(BES, aes(x=age, group = vote)) +
  geom_density(aes(fill= vote), alpha = .4) +
  geom_vline(xintercept = mean.stay, color = "blue")+
  geom_vline(xintercept = mean.leave, color = "red")
  
```

## Calculate Standard Error

Standard error: $\frac{s}{\sqrt{n}}$

```{r}
s.stay <- sd(BES$age[BES$vote=="stay"])
n.stay <- length(BES$age[BES$vote=="stay"])
se.stay <- s.stay/sqrt(n.stay)
se.stay
s.leave <- sd(BES$age[BES$vote=="leave"])
n.leave <- length(BES$age[BES$vote=="leave"])
se.leave <- s.leave/sqrt(n.leave)
se.leave
```

## Calculate Confidence Intervals

-   Ask R to formally calculate the value of t that yields a p \< 0.025 for sample sizes of 11843 and 13254 (degrees of freedom)

    ```{r}
    qt(p=.025,df = 11842)
    qt(p = .025, df = 13254)
    ```

-   CI = $\bar{x} \pm t * se$

    ```{r}
    lower.stay <- mean.stay - 1.96*se.stay
    upper.stay <- mean.stay + 1.96*se.stay
    lower.leave <- mean.leave - 1.96*se.leave
    upper.leave <- mean.leave + 1.96*se.leave
    ```

-   As a shorthand, just use t=2!

## Confidence Intervals

-   Leave: 55 $\pm$ .27 (54.73 - 55.27)

-   Stay: 47 $\pm$ .30 (46.70 - 47.30)

-   Rule of thumb: if the mean of one condition is not within the confidence interval of the other, you are able to say that the difference in the means of the two groups is statistically significant

## Hypothesis Testing

-   Null hypothesis: the means of the ages of stay and leave are the same

-   Null hypothesis: the mean of stay - the mean of leave = 0

## Test Statistic for comparing two means

-   test statistic: $$ \frac{\bar{x}_1 - \bar{x}_2}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}} $$

```{r}
t.top <- mean.stay-mean.leave
t.bottom <- sqrt(((s.stay^2)/n.stay) + ((s.leave^2)/n.leave))
t.top/t.bottom
```

-   the difference in means divided by the joint standard error (accounting for the fact that there are differences in variance and size between the samples)

## Calculate P-value

-   according to the t-distribution, what is the probability of a t-statistic less than 40?

    ```{r}
    pt(-40, 25096)
    ```

## An Easier Way!

```{r}
t.test(BES$age[BES$vote=="stay"], BES$age[BES$vote=="leave"])
```
