---
title: "W7L2: Post-Estimation Analysis"
subtitle: "POLS 642 Intermediate Analysis of Political Data"
author: "Dr. Ches Thurber"
format: 
  revealjs:
    theme: simple
execute: 
  echo: true
editor: visual
---

## Plan for Today

1\) Recap

2\) Post-Estimation Analysis

3\) Marginal Effects

4\) Predicted Values

5\) All with Interactions

## Planning Ahead

-   Thursday: Predicted Values

-   Friday 2/27: Problem Set #1 Due

-   Thur 3/5: Midterm Exam

## Packages

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)
#install.packages("lmtest")
library(lmtest)
#install.packages("sandwich")
library(sandwich)
#install.packages("modelsummary")
library(modelsummary)
#install.packages("pandoc")
#install.packages("marginaleffects")
library(marginaleffects)
gapminder <- gapminder
```

# Don't Just Crank Out Regressions!

## The Process

1\) Look at your data (make adjustments)

2\) Run your models

3\) Look at your models (make adjustments)

## We Looked at our data

```{r, echo=FALSE}
olympics <- c("United States", "Italy", "France", "Norway", "Japan", "United Kingdom", "Greece", "Australia", "China", "South Korea", "Russia", "Austria", "Switzerland")
gapminder$olympics <- ifelse(gapminder$country %in% olympics, 1, 0)
ggplot(gapminder, aes(x = gdpPercap, y = olympics))+
    geom_point() +
  scale_x_log10() +
  theme_bw()

```

## We Ran the Model

```{r, echo = F}
olymp.ols <- lm(olympics~log10(gdpPercap), data = gapminder)
modelsummary(list("LPM" = olymp.ols), output = "flextable", coef_rename = c("log10(gdpPercap)" = "Per Capita GDP (log)") ,gof_map = c("nobs", "r.squared"), vcov = ~country, stars = T)
```

## We Looked at the Model

```{r echo=F}
ggplot(gapminder, aes(x = gdpPercap, y = olympics))+
    geom_point() +
  geom_abline(slope = .16, intercept = -.49, color = "blue") +
  scale_x_log10() +
  theme_bw()

```

## We Refined The Model

```{r echo = F}
olymp.logit <- glm(olympics~log10(gdpPercap), 
                   family = binomial(link = "logit"), data = gapminder)
modelsummary(list("LPM" = olymp.ols, "Logit" = olymp.logit), output = "flextable", coef_rename = c("log10(gdpPercap)" = "Per Capita GDP (log)") ,gof_map = c("nobs", "aic") , vcov = ~country, stars = T)
```

# What sense do we make of that?!?

# Post-Estimation Analysis

## Types of Post-Estimation Analysis

-   Marginal Effects: how much does a 1-unit change in X affect our predicted value of Y?

-   Expected Values: what is our predicted value of Y across a range of values of X, holding other control variables constant

-   Many packages can help you do this; we'll use `marginaleffects()`

-   It has a website that is kinda like a textbook [here](https://marginaleffects.com)

-   I have also used `ggeffects()` , info [here](https://strengejacke.github.io/ggeffects/)

# Marginal Effects

## Average Marginal Effects

-   recall that in logit models, our regression line "curves", so the slope changes depending on the value of X

-   this is also true with log and polynomial transformations

-   average marginal effects are the average slope across the range of x

-   this often makes sense, but not always! (e.g. a u-shaped curve, or if logit curve does funny things at the extremes...)

## Calculating AME with `marginaleffects()`

```{r}
slope <- avg_slopes(olymp.logit, variables = "gdpPercap", vcov = ~country)
slope
```

## Putting it in a table with `modelsummary()`

```{r}

modelsummary(list("Logit AME" = slope), output = "flextable", 
             coef_rename = c("gdpPercap" = "Per Capita GDP") ,
             gof_map = c("nobs", "aic"), stars = T)
```

## Putting it on a plot

```{r, eval = F}
modelplot(list("Logit AME" = slope), 
          coef_rename = c("gdpPercap" = "Per Capita GDP")) +
   labs(x = 'Average Marginal Effects', 
         y = 'Variables',
         title = 'Correlates of Hosting the Olympics',
         caption = "SEs Clustered by Country") +
  theme_classic() +
  geom_vline(xintercept = 0)
```

## Putting it on a plot

```{r, echo = F}
modelplot(list("Logit AME" = slope), coef_rename = c("gdpPercap" = "Per Capita GDP")) +
   labs(x = 'Average Marginal Effects', 
         y = 'Variables',
         title = 'Correlates of Hosting the Olympics',
         caption = "SEs Clustered by Country") +
  theme_classic() +
  geom_vline(xintercept = 0)

```

# Predicted Values Plots

## Predicted Values Plots

-   this is just like when we place our regression line on our scatterplot

-   it's a little more complicated when there are multiple variables in our model

-   we might want to set those other variables to their means, or set them to specific values (e.g. oil = 1, or oil = 0)

## Using `plot_predictions()`

```{r}
plot_predictions(olymp.logit, condition= c("gdpPercap"))
```

## Fixing Up with `ggplot()`

```{r}
plot_predictions(olymp.logit, condition= c("gdpPercap")) +
  scale_x_log10() +
  theme_bw() +
  labs(x = 'Per Capita GDP', 
         y = 'Probability of Being an Olympic Host',
         title = 'Economic Development and the Olympics',
         caption = "SEs Clustered by Country") 
```

# With an interaction term

## Fitting the Model

```{r}
oil <- c("Kuwait", "Saudi Arabia", "Venezuela", "Norway", "Iran")
gapminder$oil <- factor(ifelse(gapminder$country %in% oil, 1, 0))
olymp.logit.oil <- glm(olympics~log10(gdpPercap) + oil, 
                   family = binomial(link = "logit"), data = gapminder)
olymp.logit.int <- glm(olympics~log10(gdpPercap)*oil, 
                   family = binomial(link = "logit"), data = gapminder)

```

## A Nice Table

```{r echo = F}
modelsummary(list("Base" = olymp.logit, "Oil" = olymp.logit.oil, "Interaction" = olymp.logit.int), output = "flextable", coef_rename = c("log10(gdpPercap)" = "Per Capita GDP (log)", "oil1" = "Oil Country", "log10(gdpPercap):oil1" = "GDP * Oil") ,gof_map = c("nobs", "aic") , vcov = ~country, stars = T, notes = "Std. errors clustered by country")
```

## Conditional MEs

```{r}
plot_slopes(olymp.logit.int, 
            variables = "gdpPercap", 
            condition = "oil", 
            vcov = ~country)
```

## Conditional Probabilities

```{r}
cond.probs.plot <- plot_predictions(olymp.logit.int, condition= c("gdpPercap", "oil")) +
  scale_x_log10() +
  theme_bw() +
  labs(x = 'Per Capita GDP', 
         y = 'Probability of Being an Olympic Host',
         title = 'Economic Development and the Olympics',
         caption = "SEs Clustered by Country") 
```

## The Plot

```{r echo = F}
cond.probs.plot
```
