---
title: "W4L1: Univariate Description"
subtitle: "POLS 641 Introductory Analysis of Political Data"
author: "Dr. Ches Thurber"
format: 
  revealjs:
    theme: blood
execute: 
  echo: true
editor: visual
---

## Get Started

1\) Download `W4L1_Description.zip` from Blackboard

2\) "Unzip" it and get the files set up in your file directory the way you want them

a\) `W4L1_Description.qmd`

b\) `W4L1_Description.html`

c\) `BES.csv`

3\) Open up your R Project

## What kinds of arguments do we make in political science?

-   conceptual

-   **descriptive: pure description; descriptive inference**

-   predictive

-   causal

-   normative

-   prescriptive\

## Populations and Samples

-   *Population*: all of some unit of analysis that exists (or could exist)
-   *Sample*: a subset of the population that we are able to measure (our dataset)
-   *Observation*: a single unit from our sample (row in our dataset)
-   *Variable*: measured attributes of each observation (columns in our dataset)

## Description

-   Description: characteristics of the sample

    -   Central tendency (mean, median, mode)

    -   Dispersion (quartiles, min/max, standard deviation)

## Descriptive Inference

-   Descriptive Inference: estimate the characteristics of the population based on the sample

    -   if the sample is *representative*, then it should offer an *unbiased* estimate of the characteristics of the population

    -   a sample that is *random* is often a good place to start in trying to achieve a sample that is *representative*

    -   But, in elections, this is hard...

## Potential Challenges

-   sampling frame

-   unit nonresponse

-   item nonresponse

-   misreporting

# Tables in R

## Loading Data

1\) Check your working directory

```{r}
getwd()
```

2\) Load the data (visually, or with code)

```{r}
bes <- read.csv("BES.csv")
```

## Look at the Data

Show the first five rows:

```{r}
head(bes)
```

## Creating a Frequency/Contingency Table

```{r}

table(bes$vote)
```

## Creating a Proportion Table

```{r}

prop.table(table(bes$vote))
```

# Missing Data

## What's wrong here?

```{r}

mean(bes$leave)
```

## Missing Data

```{r}

table(bes$leave, exclude = NULL)
```

## A Quick Solution

```{r}

mean(bes$leave, na.rm=T)
```

## Stop and Think!

-   I don't like the textbook's solution of deleting all missing data
-   You need to think about *why* data is missing
-   What bias might you be introducing into the inferences you make about the population when you remove missing data from your sample?
-   usually you will have to remove/ignore the missing data to proceed with analysis

# Descriptive Statistics

## Central Tendency

-   where is the "middle" of the distribution?

-   median: the "middle" observation when you put them in order

-   mean: $$\bar x = \frac{\sum_{i=1}^{n}x_i}{n}$$

## Dispersion

-   min/max

-   quintiles (value of variable at which 10%, 25%, 75%, 90% etc... of observations are lower)

-   standard deviation: like the average distance from the mean $$\sigma = \sqrt{\frac{\sum_{i=1}^n (x_i - \bar{x})^2}{n}}$$

-   variance: squared standard deviation

## Summary Stats at once

```{r}
summary(bes$age)
```

## Standard Deviation and Variance in `R`

```{r}

sd(bes$age, na.rm=T)
var(bes$age, na.rm=T)
```

# Plotting Data

## Boxplots

```{r}

boxplot(bes$age)
```

## Histogram

-   a *histogram* is a graphical representation of a variable's distribution

-   it is made of of "bins" of intervals of the variable on the x-axis (left-right)

-   the y-axis (top-bottom) represents the number of observations of the variable within each "binned" interval

## A Histogram of Respondent Age

```{r}

hist(bes$age)
```

## Density Histogram

-   a histogram where the scale of the x-axis is adjusted so that the total "area" adds up to 1

-   density of the bin = proportion of observations in the bin / width of the bin

-   it's useful because we can compare to density histograms of other samples with different numbers of observations

## A Density Histogram of Respondent Age

```{r}

hist(bes$age, freq= FALSE)
```

## Wrap up

-   Thursday: Descriptive Statistics of 2 variables
-   Next week: Visualizing Data
-   Due next Friday: Data Camp `Tidyverse` unit (four chapters)
