---
title: "W7L1: Presenting Results"
subtitle: "POLS 642 Intermediate Analysis of Political Data"
author: "Dr. Ches Thurber"
format: 
  revealjs:
    theme: simple
execute: 
  echo: true
editor: visual
---

## Plan for Today

1\) Recap

2\) Tables with Modelsummary

3\) Plots with Modelplot

4\) Visualizing SEs

## Planning Ahead

-   Today: Tables and Coefficient Plots

-   Thursday: Predicted Values

-   Friday 2/27: Problem Set #1 Due

-   Thur 3/5: Midterm Exam

## Packages

```{r}
library(gapminder)
library(dplyr)
library(ggplot2)
#install.packages("lmtest")
library(lmtest)
#install.packages("sandwich")
library(sandwich)
#install.packages("modelsummary")
library(modelsummary)
#install.packages("pandoc")
gapminder <- gapminder
```

# Don't Just Crank Out Regressions!

## The Process

1\) Look at your data (make adjustments)

2\) Run your models

3\) Look at your models (make adjustments)

## We Looked at our data

```{r, echo=FALSE}

ggplot(gapminder, aes(x = gdpPercap, y = lifeExp))+
    geom_point() +
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years",
         title = "Economic Development and Life Expectancy",
         subtitle = "Data points are country-years",
         caption = "Source: Gapminder.") + 
  theme_bw()

```

## We Made Adjustments

```{r, echo=FALSE}

ggplot(gapminder, aes(x = log10(gdpPercap), y = lifeExp))+
    geom_point() +
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years",
         title = "Economic Development and Life Expectancy",
         subtitle = "Data points are country-years",
         caption = "Source: Gapminder.") + 
  theme_bw()
```

## We Ran the Model

```{r, echo = F}
gap.model.log <- lm(lifeExp~log10(gdpPercap), data = gapminder)
summary(gap.model.log)
```

## We Looked at the Model

```{r echo=F}
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp))+
    geom_point() +
  geom_abline(slope = 19, intercept = -9, color = "blue") + 
  scale_x_log10() +
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years",
         title = "Economic Development and Life Expectancy",
         subtitle = "Data points are country-years",
         caption = "Source: Gapminder.") + 
  theme_bw()

```

## We Refined The Model

```{r echo = F}
oil <- c("Kuwait", "Saudi Arabia", "Venezuela", "Norway", "Iran")
gapminder$oil <- factor(ifelse(gapminder$country %in% oil, 1, 0))
gap.model.int <- lm(lifeExp~log10(gdpPercap)* oil, data = gapminder)
summary(gap.model.int)
```

## We Looked at it Again

```{r echo=F}
oil <- c("Kuwait", "Saudi Arabia", "Venezuela", "Norway", "Iran")
gapminder$oil <- factor(ifelse(gapminder$country %in% oil, 1, 0))
gap.model.oil <- lm(lifeExp~log10(gdpPercap) + oil, data = gapminder)
gap.model.int <- lm(lifeExp~log10(gdpPercap)* oil, data = gapminder)
ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = oil))+
    geom_point() +
  geom_abline(slope = 20, intercept = -12, color = "red") +
  geom_abline(slope = 11, intercept = 18, color = "blue") + 
  scale_x_log10() +
  labs(x = "GDP Per Capita", y = "Life Expectancy in Years",
         title = "Economic Development and Life Expectancy",
         subtitle = "Data points are country-years",
         caption = "Source: Gapminder.") + 
  theme_bw()
```

## We Looked at Diagnostic Plots

```{r, echo = F}
par(mfrow=c(2,2))
plot(gap.model.int)
```

## Plot 1: Residuals Over Fitteds

```{r, echo=F}
par(mfrow=c(1,1))
plot(gap.model.int, which = 1)
```

-   helps us evaluate all assumptions, especially linearity
-   points should be evenly distributed above and below zero

## Plot 2: QQ Plot

```{r, echo=F}
plot(gap.model.int, which = 2)
```

-   it's a quantile/quantile plot of the residuals
-   assesses the normality of the errors
-   you want it to follow closely to a straight line

## Plot 3: Scale Location Plot

```{r, echo=F}
plot(gap.model.int, which = 3)
```

-   it's a good test of homoskedasticity
-   you want a horizontal line with equally spread points
-   if not, you may have a heteroskedasticity problem

## Plot 4: Residuals vs. Leverage

```{r, echo=F}
plot(gap.model.int, which = 5)
```

-   it's a good indicator of outliers that may be skewing your results

## We tested for Heteroskedasticity

```{r}
lmtest::bptest(gap.model.int)
```

*You DON'T want to reject the null*

## We Implemented Robust Standard Errors

```{r}
coeftest(gap.model.int, vcov = vcovHC(gap.model.int, type="HC1"))
```

## We Implemented Clustered Errors

```{r}
coeftest(gap.model.int, vcov = 
           vcovCL(gap.model.int, 
                  cluster = gapminder$country))
```

# What Sense to Make of This?

## My Take

-   In this case, I think the theoretical rationale for clustered errors is so strong that I would only pay attention to models with clustered SEs

-   But there are times where multiple model specifications are justifiable and we want to show results across a range of models

-   How can we do this in a way that is clear, polished and professional?

# Presenting Regression Tables

## Some Principles for Presentation

1)  Present your findings in substantively meaningful terms

2)  Show your degree of confidence

3)  Show your data when you can

4)  Do all of this in a way that is easy for the reader to understand (?!?)

## `modelsummary()`

-   similar to stargazer

-   can do a whole lot more and more easily

-   great website [here.](https://github.com/vincentarelbundock/modelsummary)

## `modelsummary()` in practice

```{r, messages=F, warnings=F}
modelsummary(list(gap.model.log, gap.model.oil, gap.model.int), 
             vcov = ~country, stars = T, output = "flextable")
```

## Make It Look Nice

```{r, echo = F, messages=F, warnings=F}
modelsummary(list(Base = gap.model.log, Oil = gap.model.oil, Interaction = gap.model.int), coef_rename = c("log10(gdpPercap)" = "Per Capita GDP (log)", "oil1" = "Oil Country", "log10(gdpPercap):oil1" = "GDP * Oil"), vcov = ~country, gof_map = c("nobs", "adj.r.squared"), stars = T, output = "flextable")


```

## Export to Word

```{r, eval = F, messages=F, warnings=F}
modelsummary(list(Base = gap.model.log, 
                  Oil = gap.model.oil, 
                  Interaction = gap.model.int), 
             coef_rename = c("log10(gdpPercap)" = "Per Capita GDP (log)", 
                             "oil1" = "Oil Country", 
                             "log10(gdpPercap):oil1" = "GDP * Oil"), 
             vcov = ~country, gof_map = c("nobs", "R2"), 
             stars = T, 
             output = "ModelSummary_Table.docx")


```

# Coefficient Plots

## Why Coefficient Plots?

-   allows for easy visual comparisons

-   especially useful for presentations

-   "friends don't let friends present regression tables at conferences!"

## `modelplot()`

```{r, eval = F}

modelplot(list(Base = gap.model.log, 
                        Oil = gap.model.oil, 
                        Interaction = gap.model.int), 
                   coef_rename = c("log10(gdpPercap)" = "Per Capita GDP (log)", 
                                   "oil1" = "Oil Country", 
                                   "log10(gdpPercap):oil1" = "GDP * Oil"),
                   vcov = ~country) +
  theme_classic() +
  geom_vline(xintercept = 0)

```

## Coefficient Plot

```{r, echo= F}
modelplot(list(Base = gap.model.log, 
                        Oil = gap.model.oil, 
                        Interaction = gap.model.int), 
                   coef_rename = c("log10(gdpPercap)" = "Per Capita GDP (log)", 
                                   "oil1" = "Oil Country", 
                                   "log10(gdpPercap):oil1" = "GDP * Oil"),
                   vcov = ~country)
```

# Illustrating Robust SEs

## Tabular Form

```{r}

Robust.Table <- modelsummary(list(Base = gap.model.int, Robust = gap.model.int, Clustered = gap.model.int), coef_rename = c("log10(gdpPercap)" = "Per Capita GDP (log)", "oil1" = "Oil Country", "log10(gdpPercap):oil1" = "GDP * Oil"), vcov = list(vcov, vcovHC, ~country), gof_map = c("nobs", "adj.r.squared"), stars = T, output = "flextable")
```

## As a table

```{r, echo = F}
Robust.Table
```

## As a plot

```{r, echo = F}

modelplot(list(Base = gap.model.int, Robust = gap.model.int, Clustered = gap.model.int), coef_rename = c("log10(gdpPercap)" = "Per Capita GDP (log)", "oil1" = "Oil Country", "log10(gdpPercap):oil1" = "GDP * Oil"), vcov = list(vcov, vcovHC, ~country))
```

## Summary

1\) Always look at your data, both before and after running models

2\) Think carefully about which models you want to run and present to your reader

3\) Use pictures and graphs where possible

4\) Use an appendix to present additional tables of robustness checks
